Changes in mortality and child survival for mothers

### 2019-08-12 draft.Rmd
Aclarando cambio relativo. Aun por finalizar

---
title: "Changes in mortality and child survival for mothers"
author: "Ivan Williams and Diego Alburez-Gutierrez"
date: "07 August 2019"
output: pdf_document
bibliography: bib.bibtex
---

@@

The expected number of descendants for mothers is an important relation in kinship theory. Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality schedules affect the number of surviving children from the standpoint of mothers.

@@

The expected change in child survival given an absolute change in mortality depends on the difference between the age of a mother and the age structure of her surviving children. The effects of absolute changes in mortality are explained using a fertility-weighted measure of entropy.

@@

$CS_a$ is the expected number of surviving children to a woman aged $a$ in a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$:

@@

We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk, as proposed by @goodman1974, and $\delta$ can be absolute or relative.

@@

<!-- No entiendo la oracion de abajo. Le reescribi en el parrafo de arriba, pero no se si esto es lo que querias decir? -->

<!-- In other words, the magnitude of the change depends negatively on how young is the successful reproductive pattern by age, as newborns will have larger risk periods. -->

@@

Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ but weighted by the fertility pattern between [falta algo aqui?]. In this sense, depending the way in that the survival childs are accumulated by age $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$.
The effect in the child survival experienced by the mother will be bigger when the mortality is the same in all ages, no matter where the newborns are concentrated.

```{r, echo=F}
"graph"
```

@@

Considering that $m_{x,\delta}=m_{x}+\delta$ and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$:

@@

To find the effects on daughter survival of adding $\delta$ to the age-specific death rates of daughters, we get the derivative of $dCS_{(a)}^\delta / d\delta$, evaluated near zero (Keyfitz \& Caswell, 2005, section 4.3):

@@

\begin{equation}
\label{mort_change_relative}
\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{\int_{\alpha}^{a}m_{x} l_{a-x}} d\delta
\end{equation}

@@

Using that $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ is a weighting function and making an analogy with the classical entropy function $H$ where the weights are just the survival function, we can define this information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:

@@

<!-- Siento que esta parte necesita un poco mas de explicacion -->

This section proposes two additional measures of child survival for a mother aged $a$. $MTSL$ is an absolute measure of the *Mean Time Spent with Lost*, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:

\begin{equation}
\label{eq:MTSL}
MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{t,a-x}dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
\end{equation}

It is reasonable to expect a mother to relate these lost years with the expected ones, feeling the *intensity* of her losts (*Intensity Time Lost*). This makes it possible to compare between fertility regimes:

@@

Different populations could have different mean age at lost of mothers with death child:

@@

The proposed meassures are applied to two different demographic regimes.

@@

A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.

>>>>

---
title: "Changes in mortality and child survival for mothers"
author: "Ivan Williams and Diego Alburez-Gutierrez"
date: "07 August 2019"
output: html_document
bibliography: bib.bibtex
---

@@

The expected number of descendants for mothers is an important relation in kinship theory. Child survival (and child death) affect mothers in a variety of ways, yet it is not clear how changes in mortality schedules, given a fertility pattern, affect the number of surviving children from the standpoint of mothers.

@@

The expected change in child survival given an absolute change in mortality depends on the difference between the age of a mother and the age structure of her surviving children. The effects of relative changes in mortality are explained using a fertility-weighted measure of entropy.

@@

$CS_a$ is the expected number of surviving children to a woman aged $a$ in a stable female population with fertility rates $m_{x}$ and force of mortality $\mu_{x}$, as proposed by @goodman1974:

@@

We now consider the consequences of a change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.

@@

@@

Where $H_{CS_a}$ can be interpreted as a temporary entropy constant until age $a-\alpha$ . 
Depending the age structure of the survival daughters $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$,  is the way that is taken each age contribution to the disparity measure $H_{CS_a}=\int_\alpha^a{C_{CS_a}(x)H(x)}$ (@Aburto2019).

The effect of a relative change $\delta$ will be bigger when the mortality is more disperse in ages where the fertility is very concentrated. For example in \ref{fig:relative_change}, the youngest schedule would have a bigger change in $C_{CS_a}$...

```{r relative_change, echo=F, warning=F, fig.cap="Two different fertility schedules (black lines) and a big hump mortality case (right axis, line in red). In dashed grey lines are the mean age at birth"}

# genero m_x (old and young pattern) y l_x (with big hump) 
f_young = data.frame(x=c(15,20,25,30,35,40), f=c(0,.3,.3,.05,.01,.01))
f_old = data.frame(  x=c(15,20,25,30,35,40), f=c(0,.04,.06,.25,.1,.01))
f_young_p = predict(loess(f~x, data = f_young), 15:40)
f_old_p = predict(loess(f~x, data = f_old), 15:40)
f_young_mean = sum(f_young_p*15:40/sum(f_young_p))
f_old_mean = sum(f_old_p*15:40/sum(f_old_p))
m_si_hump = data.frame(x= c(0,1,5,10,15,20,25), l=c(1,.98,.97,.97,.96,.9,.85))
m_si_hump_sp = splinefun(m_si_hump, method = "monoH.FC")

#graph
library(latex2exp)
par(mfrow=c(1,1),mar = c(5,5,2,5))
plot(15:40, f_young_p, t="l",xlab="x (mother age)", ylab= TeX(sprintf("$m_x$"))) 
lines(15:40, f_old_p, lty=2) 
abline(v=f_young_mean, lty=2, col="grey")
abline(v=f_old_mean, lty=2, col="grey")
par(new = T)
plot(0:25, rev(m_si_hump_sp(0:25)), axes=F, xlab=NA, ylab=NA, t="l", col=2)
axis(side = 4)
mtext(side = 4, line = 3, TeX(sprintf("$l_{40-x}$")))

#calculate entropy CS
H_young = -sum(f_young_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
  sum(f_young_p*rev(m_si_hump_sp(0:25)))
H_old = -sum(f_old_p*rev(m_si_hump_sp(0:25))*log(rev(m_si_hump_sp(0:25))))/
  sum(f_old_p*rev(m_si_hump_sp(0:25)))
```

@@

Considering that $m_{x,\delta}=m_{x}+\delta$ ((@wrycza2012)), and $l_{a-x} = e^{-\int_{0}^{a-x}{(\mu_t+\delta})}$ :

@@

To find the effects on daughter survival of adding $\delta$ to the age-specific death rates of daughters, we get the derivative of $dCS_{(a)}^\delta / d\delta$, evaluated near zero (@KeyfitzMath, section 4.3):

@@

\begin{equation}
\label{mort_change_relative}
\frac{d CS_{a}}{CS_{a}} = \frac{ \int_{\alpha}^{a}m_{x} l_{a-x} \log(l_{a-x}) dx}{CS_{a}} d\delta
\end{equation}

@@

Calling $C_{CS_a}(x)=\frac{m_{x}l_{a-x}}{CS_{a}}$ the weighting function of the succesful fertility pattern (onset distribution of survival daughters), and making an analogy with the classical entropy function $H$ where the weights are just the survival function (@KeyfitzMath), we can define this information function $H_{CS_a}=\int_\alpha^a{C_{CS_a}H(x)}$, and get:

@@

This section proposes three additional measures related to the child survival measure for a mother aged $a$.

*Mean Time Spent with Lost* ($MTSL$) is an absolute measure of the expected time with a death daughter that lives a mother, which can be expressed in terms of a temporary expected lost years measure $e^\dagger$:

\begin{equation}
\label{eq:MTSL}
MTSL_a = \int_\alpha^a{m_x\int_0^{a-x}{d_t e_{0,a-x-t} dt}dx} = \int_\alpha^a{m_x e_{o,a-x}^\dagger dx}.
\end{equation}

Where $d_t$ is the death distribution from birth, and $e_{0,a-x-t}$ is the life expectancy at birth until age $a-x-t$.

It is reasonable to think that mothers relates these lost years with the expected time with live daughters, getting the *intensity* of her losts (*Intensity Time Lost* ($ITL$)). This makes possible also to compare between fertility regimes:

@@

Different populations could have different mean age at lost ($MAL$) of mothers with death child:

@@

<!-- Aquí tenemos dos opciones de aplicación  -->

* The proposed meassures are applied to two different demographic regimes.

@@

* A computation approach is used to evaluate the relative contribution of each component given a relative change in $\delta$.
